Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-04-25 paper.tex
edits mainly to section 2.  Please proofread and check.  It should be obvious but I mainly wanted to (aO cut ideas already in the intro (2) expand on the Richardson like argument (i.e not force the reader to go to Urpin to follow the flow) (c) put the figs a bit earlier (e.g. fig 1 where we first give shear and buoyancy) and (d) expand a bit the physical discussion of t_c figure and implications, but I could be convinced that some of this belongs more in the concluding discussion (on this is was just worried, maybe too much, about giving some wrong impressions).

\section{Thermodynamic requirement for the vertical shear instability}\label{vsi_require}     In this paper we will find that in a stably stratified thin disk, the VSI  
requires a thermal timescale $t_c$ much shorter than the
disk dynamical timescale, 
\begin{align}
  t _c \ll \OmK^{-1}. 
\end{align}
where $\OmK$ is the Keplerian frequency, consistent with N13. 
To see why this should be expected, we need to compare vertical shear
with vertical buoyancy. For this preliminary discussion we consider a
vertically isothermal disk with aspect-ratio $ h$,  
dimensionless radial temperature gradient $q$ and adiabatic index
$\gamma$. For a PPD, $ h \sim 0.05$, $|q|\sim 1$ and $\gamma\simeq 1.4$.

For a thin disk, the specific angular momentum varies with height as 
\begin{align}\label{vshear_thin}
  r \frac{\p \Omega^2 }{\p z} \simeq  h q \OmK^2\left(\frac{z}{H}\right),
\end{align}
where $H$ is the disk scale height. As noted in
\S\ref{intro} this vertical variation is a source of free energy and
thus vertical shear ($\p_z\Omega \neq0$) is destabilizing.

For $\gamma>1$ the disk is stably stratified with a real vertical
buoyancy frequency $N_z$.  In a thin disk,     
\begin{align}\label{nz_thin}
  N_z^2 \simeq \left(\frac{\gamma-1}{\gamma}\right) \left(\frac{z}{H}\right)^2
  \OmK^2.  
\end{align}
This provides a stabilizing effect as buoyancy opposes vertical
motion, which is required for the VSI. In fact, if we consider
axisymmetric adiabatic disturbances, then because the 
Solberg-Hoiland criteria \citep{tassoul78} are usually satisfied in a
thin disk, the disk would be stable.

However, the stabilizing effect of buoyancy can be mitigated for 
effectively isothermal disturbances. Consider a fluid element displaced upwards, away from
the disk midplane. In a stably stratified adiabatic disk, it would
sink because of buoyancy. However, if thermal timescales are short, in
that temperature flucations decay rapidly, then the fluid element would attain its surrounding 
density and  
%(assuming an ideal gas and it reaches pressure 
%equilibrium with its surroundings). 
thus buoyancy is rendered ineffective. Qualitatively, a short thermal
timescale $t_c$ reduces the effective (stabilizing) buoyancy force.

How short must $t_c$ be for vertical shear to prevail? A rough 
comparison of vertical shear to buoyancy for non-adiabatic
perturbations can be obtained by applying Eq. 47 of \cite{urpin03}, which is based on 
\cite{townsend58},   
\begin{align}\label{bcrit_est}
  N_z^2 t_c \lesssim \left|r\frac{\p\Omega}{\p z}\right|. 
\end{align}
The left-hand side represents an effective buoyancy, which is
stabilizing; while the right-hand side represents vertical shear,
which is destabilizing. In a thin disk $ h \ll 1$, so the destabilizing effect of 
vertical shear is rather weak because $|r\p_z\Omega|\sim h\OmK
\ll \OmK$. By comparison, the buoyancy frequency $N_z\sim
\OmK$. It follows that $t_c\lesssim O( h\OmK^{-1})$ is
required for instability.  \emph{In a thin  
  disk the thermal timescale must be shorter than the dynamical
  timescale by $O( h)\ll 1$ for the VSI to operate.}

To illustrate this point, we show in Fig. \ref{eqm_structure} the
equilibrium structure of a typical PPD. %Here $\rho_0$
%is the midplane density and vertical shear rate and buoyancy
%frequencies are normalized by the local Keplerian frequency. 
It is
clear that vertical shear is 
generally weak compared to buoyancy. Then even a small (positive)
vertical entropy gradient can significantly stabilize the VSI. 
Note that in astrophysical disks both $N_z^2$ and $\p_z\Omega$ varies
with height. Eq. \ref{bcrit_est} then suggest that with
increasing height, buoyancy would eventually stabilize the VSI (since
$N_z^2\propto z^2$ and $\p_z\Omega\propto z$).
% This means that 
% disturbances localized away from the disk midplane are easily stabilized by
% buoyancy. 
This argument assumes vertically localized disturbances. It is,
however, not clear how to apply Eq. \ref{bcrit_est} to vertically
global disturbances, as considered in this work.

\begin{figure*}
  \includegraphics[width=\linewidth]{figures/rhoshearNz}
  \caption{Typical PPD structure (the Minimum Mass
    Solar Nebla): density (left), vertical shear rate (middle) and
    buoyancy frequency (right). Here $\rho_0$ is the midplane density.  
    \label{eqm_structure} 
  }
\end{figure*}

Our central result is the requirement 
\begin{align}\label{prelim_bcrit}
  (\gamma - 1)t_c <  h |q| \OmK^{-1}
\end{align}   
for the VSI in a vertically isothermal, thin disk. This equation
reflects Eq. \ref{bcrit_est} but there is no $z$ dependence. 
% While assumptions and simplifications will be made 
% to derive this, 
Eq. \ref{prelim_bcrit} can be 
readily applied to evaluate where in a disk the VSI is likely to
operate. For example, we show in Fig. \ref{bcrit_mmsn2d} an estimate
of the thermal timescales the PPD model shown in
Fig. \ref{eqm_structure}. Here $k_x$ is a radial
wavenumber of the disturbance being considered, thus larger $k_x$
correspond to smaller perturbations. The region outside the red line
satisfies Eq. \ref{prelim_bcrit}.

The criterion Eq. \ref{prelim_bcrit} in fact
assumes $t_c$ is vertically uniform. While this is not generally true
in the inner disk, we find Eq. \ref{prelim_bcrit} still provides a
useful guide. We see that the VSI will operate in the outer 
disk at tens of AU, but that smaller scale disturbances may operate at
smaller radii.

@@

We now proceed through the formal linear stability problem to arrive
at Eq. \ref{prelim_bcrit} and verify it numerically. The interested 
reader may skip forward \S\ref{application} for the application of our
results to PPD.

>>>>

\section{Thermodynamic requirement for the vertical shear instability}\label{vsi_require}     
In a stably stratified thin disk, the VSI  
requires a thermal timescale $t_c$ much shorter than the
disk dynamical timescale, 
\begin{align}
  t _c \ll \OmK^{-1}. 
\end{align}
where $\OmK$ is the Keplerian frequency (N13).    
This requirement -- that the cooling time be much shorter than an orbital period, 
which in turn is much smaller than the relevant oscillation period of vertically elongated gravity waves  -- is quite stringent.
It highlights the fact that vertical buoyancy is a very strong stabilizing influence.
Very strong radiative damping is thus required to weaken buoyancy and allow the weak vertical shear to drive instability.

To roughly quantify the required smallness of $t_c$, we consider a
vertically isothermal disk with aspect-ratio $ h$, 
 radial temperature profile $T \propto r^q$ and adiabatic index
$\gamma$. For a PPD, $ h \sim 0.05$, $q\sim -0.5$ and $\gamma\simeq 1.4$.

\begin{figure*}
  \includegraphics[width=\linewidth]{figures/rhoshearNz}
  \caption{The radial and vertical structure of a ``minimum mass" PPD model showing gas density (left, relative to the midplane density at 1 AU), vertical shear rate (middle) and
    buoyancy frequency (right).
    \label{eqm_structure} 
  }
\end{figure*}

For a thin disk, the vertical shear increases with height, $z$, from the midplane as 
\begin{align}\label{vshear_thin}
  r \frac{\p \Omega }{\p z} \simeq  h q \OmK \left(\frac{z}{2 H}\right),
\end{align}
for several disk scale heights, $H = hr$. 
%As noted in
%\S\ref{intro} this vertical variation is a source of free energy and
%thus vertical shear ($\p_z\Omega \neq0$) is destabilizing. 
%
%For $\gamma>1$ the disk is stably stratified with a real 
This destabilizing shear completes with the stabilizing vertical
buoyancy frequency $N_z$.  In a thin disk,     
\begin{align}\label{nz_thin}
  N_z^2 \simeq \left(\frac{\gamma-1}{\gamma}\right) \left(\frac{z}{H}\right)^2
  \OmK^2.  
\end{align}
Fig. \ref{eqm_structure} maps shear and buoyancy (as well as the gas density) 
in a fiducial PPD model, whose parameters are in Appendix \ref{mmsn} . %Here $\rho_0$
%is the midplane density and vertical shear rate and buoyancy
%frequencies are normalized by the local Keplerian frequency. 
Vertical shear is 
generally weak compared to buoyancy, suggesting stability.

Without any cooling, the Solberg-Hoiland criteria confirms that vertical buoyancy is strong
enough to ensure (axisymmetric) stability, see \S\ref{solberg}.  With radiative cooling, 
thermal fluctuations decay which, combined with pressure equilibration, reduces the effective buoyancy.

%However, the stabilizing effect of buoyancy can be mitigated for 
%effectively isothermal disturbances. Consider a fluid element displaced upwards, away from
%the disk midplane. In a stably stratified adiabatic disk, it would
%sink because of buoyancy. However, if thermal timescales are short, in
%that temperature flucations decay rapidly, then the fluid element would attain its surrounding 
%density and  
%%(assuming an ideal gas and it reaches pressure 
%%equilibrium with its surroundings). 
%thus buoyancy is rendered ineffective. Qualitatively, a short thermal
%timescale $t_c$ reduces the effective (stabilizing) buoyancy force.

How short must $t_c$ be for vertical shear to prevail?  We start by 
considering the Richardson number $\mathrm{Ri} = N_z^2/(r\p_z\Omega)^2$, a 
ratio of buoyant to shear energies.  Though not precisely applicable to our problem,
non-rotating shear flows are stable if $\mathrm{Ri} > 1/4$ (\citealp{chandrasekhar61}, 
also see \citealp{ys02, lee10} for applications to thin dust layers in PPDs).
Following \cite{urpin03} and \cite{townsend58},  we reduce the buoyant energy by the 
ratio of cooling to forcing timescales, $t_c |r \p_z\Omega| \leq 1$ (with no reduction in buoyant energy 
expected when this inequality fails).  With this reduction, the Richardson like criterion for 
shear \emph{instability} becomes
\begin{align}\label{bcrit_est}
  t_c \lesssim \frac{\left|r\p\Omega/\p z\right|} {N_z^2}. 
\end{align}

This result is far from rigorous, but it contains some useful insights.
 If we interpret Eq.\ \ref{bcrit_est} as a local criterion, then we see that 
 the minimum cooling time decreases with height as $t_c \propto 1/z$.
 This finding is relevant to the radiative damping of so called ``surface 
 modes," which we consider in \S\ref{numerical}.

 To roughly obtain a global cooling time criterion, we evaluate Eq.\ \ref{bcrit_est}
 at $z \sim H$.  To get the precise form we want, we choose $z = \gamma H / 2$ to 
 obtain the cooling requirement
 \begin{align}\label{prelim_bcrit}
 t_c <  h |q| \OmK^{-1}/ (\gamma - 1)
\end{align}   
for the VSI in a vertically isothermal, thin disk.  This estimate agrees with 
our central result, the cooling criterion that we derive more rigorously in \S\ref{iso_vsi_beta_crit}.

%
%The left-hand side represents an effective buoyancy, which is
%stabilizing; while the right-hand side represents vertical shear,
%which is destabilizing. In a thin disk $ h \ll 1$, so the destabilizing effect of 
%vertical shear is rather weak because $|r\p_z\Omega|\sim h\OmK
%\ll \OmK$. By comparison, the buoyancy frequency $N_z\sim
%\OmK$. It follows that $t_c\lesssim O( h\OmK^{-1})$ is
%required for instability.  \emph{In a thin  
%  disk the thermal timescale must be shorter than the dynamical
%  timescale by $O( h)\ll 1$ for the VSI to operate.}   
%
%
%Then even a small (positive)
%vertical entropy gradient can significantly stabilize the VSI. 
%Note that in astrophysical disks both $N_z^2$ and $\p_z\Omega$ varies
%with height. Eq. \ref{bcrit_est} then suggest that with
%increasing height, buoyancy would eventually stabilize the VSI (since
%$N_z^2\propto z^2$ and $\p_z\Omega\propto z$).
%% This means that 
%% disturbances localized away from the disk midplane are easily stabilized by
%% buoyancy. 
%This argument assumes vertically localized disturbances. It is,
%however, not clear how to apply Eq. \ref{bcrit_est} to vertically
%global disturbances, as considered in this work.
%
%
%
%
%Our central result is the requirement 
%\begin{align}\label{prelim_bcrit}
%  (\gamma - 1)t_c <  h |q| \OmK^{-1}
%\end{align}   
%for the VSI in a vertically isothermal, thin disk. This equation
%reflects Eq. \ref{bcrit_est} but there is no $z$ dependence. 
%% While assumptions and simplifications will be made 
%% to derive this, 
%Eq. \ref{prelim_bcrit} can be 
%readily applied to evaluate where in a disk the VSI is likely to
%operate.

@@

We now consider whether, and where, thermal timescales in realistic PPDs
are short enough to support VSI growth.  Fig. \ref{bcrit_mmsn2d} plots the
 thermal timescales in our fiducial PPD model with a small grain dust opacity. 
 Since cooling by radiative diffusion depends on a perturbation's lengthscale, we
 plot $t_c$ for two values of the radial wavenumber,  $k_x$,  the relevant 
 component for VSI modes that are radially narrow.   The red line
marks the boundary implied by Eq. \ref{prelim_bcrit}.  We caution that 
 Eq. \ref{prelim_bcrit} applies when $t_c$ is vertically uniform.  This section is 
 aimed at basic insights, while \S\ref{application}
 analyzes the VSI with realistic spatial variations in $t_c$.

Due to the smaller values of $t_c$ in the outer disk, we generally expect the VSI to 
operate outside $\sim 10$ AU.  This conclusion changes however, if the dust opacity is reduced, 
slowing down cooling, as we show later (e.g.\ Fig.\ \ref{mmsn_bcrit_bcool}).

In the inner disk we see that thermal times are long in the midplane, but short at higher altitutdes. 
Again, this structure necessitates the more detailed analyses that follow.  Nevertheless, we can already 
see that due to radiative diffusion higher wavenumbers can cool more rapidly.  Thus smaller scale disturbances 
may trigger VSI at smaller disk radii.

However we cannot and do not find that VSI will 
always be relevant for sufficiently small perturbations.  For starters, very small wavelengths are in the optically 
thin cooling regime,  where cooling is independent of lengthscale.  Moreover, small scale perturbations 
are prone to viscous damping, and should drive weaker transport.

We now proceed with our detailed analysis of the VSI to arrive
at Eq. \ref{prelim_bcrit} and verify it numerically. The interested 
reader may jump to \S\ref{application} for the application of our
results to PPDs.
